---
title: "Analysising the Gender Pay Gap In The U.S Workplace"
author: ""
date: "2020/12/22"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Setups & Preparation
library(tidyverse)
library(lme4)
library(table1)
final_data <- read.csv("E:\\Assignments\\STA304\\Sta304-Final Project\\Sta304 - Final Project cleaned data.csv")

propensity_score <- glm(treatment ~ age + wkswork1 + educd + stateicp + classwkrd + race,
                      family = binomial,
                      data = final_data)

final_data_prob <- 
  augment(propensity_score, 
          data = final_data,
          type.predict = "response") %>% 
  dplyr::select(-.resid, -.std.resid, -.hat, -.sigma, -.cooksd) 

matches <- arm::matching(z = final_data_prob$treatment, 
                         score = final_data_prob$.fitted)
final_data_prob <- cbind(final_data_prob, matches)

final_data_matched <- 
  final_data_prob %>% 
  filter(match.ind != 0) %>% 
  dplyr::select(-match.ind, -pairs)

propensity_score_regression <- 
  lm(log(inctot) ~ age + race + educd + classwkrd + stateicp + wkswork1 + sex, 
                data = final_data_matched)
```


# Analysising the Gender Pay Gap In The U.S Workplace

## Erdong Zhang
## 2020/12/22
## Github repo link:

# Abstract
  The gender pay discrepancy is a social problem that is shown as a gender imbalance in the workplace’s payments. This study applies the approach of propensity score matching and suits a linear regression model to find out the difference of gender income in the U.S. workplace. The result is significant and reveals that holding age, race, state of residence, education attainment, hours of work, and worker fixed class. The man, on average, continues to receive a 37.3% higher annual pre-tax income than the female. The outcome shows that gender wage inequality is sexism against women in the workplace and the implications of the disparity in different ways impact the U.S. economy and the lives of women.
  
## Keywords 
Gender pay gap, propensity score matching, Discrimination, Disparity

# Introduction
  The U.S used to be a pioneer in tackling gender income inequality. Of the rich countries, it was the first country to pass laws restricting occupational gender inequality. However, it currently has a more significant gender pay gap than most OECD countries (Glynn, 2018). Additionally, Glynn (2018) notes that the prevalence of wage disparity between men and women would directly and implicitly impact the U.S economy. In addition, she suggests that the gender wage gap decreases family wealth, weakens spending power, and directly impacts single-parent families with low and medium incomes.
  Moreover, Farrell and Glynn (2013) identify many factors. For instance, differences in work experiences, occupations, and industries could explain a portion of the wage disparity. They stress, though that nearly half of the reasons related to gender wage inequality are not measurable, and may be overt bias or unconscious prejudice against women that prohibits them from bargaining in the workplace for a better payment. Hence, drawing a causal inference in this context is the key to uncover the unobservables and discover the impact of gender discrimination on wage disparities in the U.S workplace.
  One way to determine the causal inference is the propensity score matching method. Propensity score matching is a popular, powerful, and robust method for determining the causal effect on observational data (Arbour et al., 2014). This report adopts the propensity score matching method. It complies with the latest U.S microdata that aims to justify that gender discrimination is pervasive in the U.S current work environment and causes vast income disparity between men and women. Since the last century, the U.S government has implemented laws and regulations to address the gender pay gap issue and improve the corresponding legislation. However, the pay gap has not been significantly narrowed so far. According to Bleiweis (2020), “the gender pay gap has only closed by 4 cents for every $1 earned by men in more than a decade.” Also, she points out that absent any reforms in current legislations, closing the gender pay gap is infeasible in the next forty years. Thus, by precisely specify the causal effect between gender discrimination and the gender pay gap, this report also suggests that the U.S government should take further measures to eliminate gender discrimination and protect females’ equality in the workplace.
  The following Methodology section contains Data and Model subsections. The Data section provides information regarding data collection and a baseline characteristic table of the data. The model section shows the specific equations of the model and discusses relevant features that enter the model. The result of the propensity score analysis is provided in the Result section. Interpretations of the result and further discussions related to the gender pay gap are in the Discussion section.

# Methodology

## Data
  This report uses the 2019 American Community Survey (ACS) dataset retrieved from Integrated Public Use Microdata Series (IPUMS-USA). IPUMS-USA provides easily accessible U.S census microdata that consists of abundant demographic and economic variables. According to the ACS information guide (United States Census Bureau, 2017), the data is collected through the following approaches: The United States Census Bureau randomly selects household addresses across the country with each has 1/480 probability of been determined, and the same household should not be chosen more than once every five years. The majority of the data is collected via mailing the letters to the randomly selected addresses that invite people living in the address to participate in an online survey. If the Census Bureau does not receive a completed response in a few weeks, the Census Bureau will send an additional paper questionnaire to the address. Besides, the Census Bureau will conduct a personal interview survey for people living in group housing such as nursing homes, prisons, and college dormitories. To deal with non-respondents or uncompleted surveys, the Census Bureau (1) sends a field representative to conduct a personal interview with the address that did not respond to the survey online or through the mail. (2) conducts a telephone follow-up.
  The data used in this report contains 31437 randomly selected observations from the 2019 ACS dataset. It has eight variables, which are age, sex, race, educational attainment, state of residence, class of worker, weeks of working last year, and annual pre-tax personal income (will be referred to as personal income for simplicity), where sex, race, educational attainment, state of residence, and a class of worker are categorical variables; age, weeks of working, and personal income are numerical variables. A new variable called treatment is created to determine the propensity score to perform the propensity score matching method to achieve the propensity score matching method. The variable treatment is a dummy variable with a value of 1 if the sex is female and equal to 0 if the sex is male. Table 1 provides a detailed baseline characteristics table of the data. It shows the mean and median values of age, personal income, and weeks of working last year for males (control group) and females (treated group). It also displays every category for education attainment, state of residence, race, class of worker, and the composition of the male and female in each category. A notable finding in Table 1 is that there is a clear difference in the average personal income between male and female given their average weeks of working and ages are nearly the same, which means female in general earn less than the male of the same period under the same amount of work, suggesting gender income disparity does exist in the U.S. workplace. 

## Model
  For the purpose of testing the hypothesis that there is a gender pay gap in the U.S. workplace, this report incorporates a logistic regression model and multiple linear regression model in computing the propensity score and ascertaining the gender effect on wages, respectively. Both models are run in the R-studio, more specifically, The logistic regression model to calculate the propensity score reads as:
$$ log(\frac{p}{1-p})_i = \tilde\beta_0 + \tilde\beta_1 age_{i} + \tilde\beta_2{wkswork1}_i + \tilde\beta_3{educd}_i+ \tilde\beta_4{stateicp}_i + \tilde\beta_5{classwkrd}_i + \tilde\beta_6{race}_i +\epsilon_i$$
  where $p$ represents the probability of being assigned to the treatment group for each I in the sample. The estimated coefficient $\tilde\beta_1$ captures the change in log odds for one year increasing in age and $\tilde\beta_2$ represents the change in the log odds for each additional week of working. Besides, $\tilde\beta_3$ to $\tilde\beta_6$ represent the effect of a specific category on changes in log odds. Lastly, $\epsilon$ is the error term.
  Since the dependent variable, treatment is a binary dummy variable. The logistic regression model will provide a more precise estimate for a binary dependent variable than other models such as linear regression models. It is the main reason for choosing the logistic regression model in determining the propensity score. 
  After finishing calculating the propensity score for each individual in the sample and the pair matching process, this report proposes a multiple linear regression model to estimate the gender income disparity. The model has the following format:
$$ log(inctot_i) = \beta_0 + \beta_1 {age}_i + \beta_2 {race}_i + \beta_3 {educd}_i + \beta_4 {stateicp}_i + \beta_5 {classwkrd}_i + \beta_6 {sex}_i + u_i $$
  where $\beta_1$ represents the change in pre-tax annual personal income for the individual i in the sample as getting one year older. $\beta_2$ to $\beta_5$ specifies the magnitude effect of demographic and geographic factors on the pre-tax annual personal income for the individual i. $\beta_6$ is the primary interest of the study in this report, which estimates the gender pay gap, and $u_i$ is the error term of the linear regression.
  The model is well predicted and fits the data since it has an R-squared value of 0.529, and model diagnostic plots (See Figure 1) do not violate linear assumptions. Significantly, residuals are randomly spread out around a horizontal line without a distinct pattern, as shown in the residuals vs. fitted plot. Besides, the typical Q-Q plot indicates the assumption of residual normality is reasonable. Additionally, in the scale-location plot, dots are randomly equal spread out along the asymptotic horizontal line with a few outliers, suggesting the homoscedasticity of residual. Lastly, all the points inside the Cook’s distance, including outliers in the residual vs. leverage plot, there is no influential sample that affects the model prediction. Overall, the multiple linear regression model is adequate for estimating and feasible determining the gender pay gap.
  Moreover, the report chooses to use the logarithm of personal income (log(inctot)) as the dependent variable instead of personal income. The log income’s estimated coefficients can provide a more explicitly fraction form result to show that females tend to earn $\beta_6$ cents less for every $1 made by a male on average, holding age, race, educational attainment, state of residence, and a class of worker fixed.

# Result
Statistical results of the multivariate  analysis is presented below.
```{r}
huxtable::huxreg(propensity_score_regression)
```

The model’s estimate implies that assuming age, race, educational attainment, state of residence, class of worker, and weeks of working are equal, females on average tend to earn 37.3% less pre-tax annual income than males in the U.S workplace. The difference is statistically significant in a 1% significance level and also economically significant as females, in general, earn 37.3 cents less for every 1 dollar made by males under Ceteris paribus. The result is based on the propensity score matching method that applies the logistic regression to determine the propensity score and uses multivariate analysis based on the new sample that has been processed by the propensity score matching method. Besides, the estimate $\beta_6$ has a p-value less than 0.01 implies the data provides sufficient pieces of evidence to conclude the gender pay gap does exist in the U.S. workplace

# Discussion
  The gender pay gap is a long-standing phenomenon that pervades the workplace in the United States, and since the last century, the appeal for wage equality has not ceased. Many factors cause the gender pay gap; this report interests the gender discrimination effect on income. Moreover, being able to derive a causal inference is essential for an analytical study. To establish the causal association between gender and income inequality, this study adopts the likelihood score matching process. Also, the data used in this report provide a significant advantage in deriving the causality as households are randomly selected by the Census bureau. Lastly, the information uses logistic regression to determine the propensity score and propose a multiple linear regression model to estimate the gender pay gap’s magnitude.
  According to the propensity score analysis, holding age, race, education attainment, class of worker, and weeks of working the same, females on average earn 37.3% less than males in the U.S. workplace. It is a huge income gap in practice and has a vital statistical significance. In other words, the result justifies that overt sexism causes females to earn 37.3 cents less for every $1 made by males. Therefore while legislation has been adopted by the U.S. government to reduce the gender wage differential created by gender inequality against women, the statistical outcome suggests that the U.S. government should take more steps to mitigate the impact of misogyny on wages.
  There may however be concerns of homogeneity in the sample that could theoretically influence the organization and the accuracy of calculating the gender wage difference. Due to privacy issues or other purposes, because much of the data is gathered from online surveys, participants could prefer to distort their personal details, especially income. Increasing the sample size would reduce bias due to misleading responses.  Future studies could consider including interaction terms or analyzing the race income disparity.





# Appendix
Table 1: Baseline Characteristics Table
```{r}
# Baseline characteristics of the data
Table <- table1(~ + age + inctot + wkswork1+ educd + stateicp + race + classwkrd | sex , data = final_data)
Table

```


Figure 1: Model Diagnostic plots
```{r}
# Diagnostic plots
par(mfrow=c(2,2))
plot(propensity_score_regression)
```

# Reference
Arbour, D., Marazopoulou, K., Garant, D., & Jensen, D. (n.d.). Propensity Score Matching for Causal Inference with Relational Data. [PDF]. University of Massachusetts Amherst. https://staff.fnwi.uva.nl/j.m.mooij/uai2014-causality-workshop/papers/paper5.pdf. 

Bates, D., Maechler, M., Bolker, B., Walker, S. (2015). Fitting Linear Mixed-Effects Models
Using lme4. Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.

Bleiweis, R. (2020, March 24). Quick Facts About the Gender Wage Gap.https://www.americanprogress.org/issues/women/reports/2020/03/24/482141/quick-facts-gender-wage-gap/.

Farrell, J., & Glynn, S. J. (2014, July 8). What Causes the Gender Wage Gap? Center for American Progress. https://www.americanprogress.org/issues/economy/news/2013/04/09/59658/what-causes-the-gender-wage-gap/.

Glynn, S. J. (2019, September 25). Gender wage inequality. Equitable Growth. https://equitablegrowth.org/research-paper/gender-wage-inequality/?longform=true.

Rich, B. (2020). table1: Tables of Descriptive Statistics in HTML. R package version 1.2.1.
https://CRAN.R-project.org/package=table1

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.


Ruggles, S., Flood, S., Goeken, R., Grover, J., Meyer, E., Pacas, J., & Sobek, M. (2020). IPUMS USA: Version 10.0 [Data set]. Minneapolis, MN: IPUMS. https://doi.org/10.18128/D010.V10.0

Understanding diagnostic plots for linear regression analysis | university of virginia library research data services + sciences. Retrieved from https://data.library.virginia.edu/diagnostic-plots/

United States Census Bureau (2017, October). American Community Survey Information Guide. [PDF]. U.S Department of Commerce. 

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686
